Automatic word stress annotation of Russian running text (max 8 pages)

### 2015-01-21 rusaccent-nodalida-2015/rusaccent.tex
Added more citations in bibtex and cleaned up footnotes. Formatting is now submittable. Need to remove marginpars.
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58514 72bbbca6-d526-0410-a7d9-f06f51895060

\subsection{Background and Task Definition} \label{background}

@@

\subsection{Stress Corpus}

@@

\section{Automatic Stress Placement}

State-of-the-art morphological analysis in Russian is based on finite-state
technology.\footnote{add footnote or citations for Dialog, mystem, etc.} To our 
knowledge, no open-source resources are available for analyzing and generating 
stressed wordforms. Therefore, we developed free and open-source finite-state 
tools capable of analyzing and generating stressed wordforms, based on the
well-known \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
A Finite-State Transducer\footnote{Using two-level 
morphology, xfst/hfst, blah, blah, blah} (FST) generates all possible 
morphosyntactic readings of each wordform, and a constraint grammar\footnote{vislcg3, blah blah} then removes
some extra readings based on syntactic context. The ultimate success of our 
stress placement system depends on the performance of the Constraint Grammar. 
Ideally, the Constraint Grammar would successfully remove all but the correct 
reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

@@

The {\small {\tt guessFreq}} 
approach is to select the reading that is most frequent, with frequency data
taken from a separate hand-disambiguated corpus. If none of the readings is 
found in the corpus, then {\small {\tt guessFreq}} selects the reading
with the tag sequence (lemma removed) that is most frequent in our corpus. If the 
tag sequence is not found
in our frequency list, then {\small {\tt guessFreq}} backs off to the 
{\small {\tt guess}} algorithm. In the first column of
Table~\ref{tab:conditions}, {\small {\tt guessFreq}} selects \rus{к\'{о}сти} 
\emph{k\'{o}sti} because the tag sequence N-F-SG-GEN is more frequent than the
other alternatives. Note 
that for tokens with stress-irrelevant ambiguity (e.g. the second column of
Table~\ref{tab:conditions}), {\small {\tt guess}} and 
{\small {\tt guessFreq}} produce the same result as the {\small {\tt safe}} 
method.

@@

We noted in Section~\ref{background} that \newcite{hall_sproat_russian_2013}
achieved 98.7\% accuracy on stress placement for individual wordforms 
in a list (i.e. \emph{without} syntax). This result is 8.63\% higher than
{\small {\tt noCGsafe}}, but it is also a fundamentally different task.
Based on the surface 
forms in our FST -- which is based on the same dictionary used for 
\newcite{hall_sproat_russian_2013} -- 
we calculate that only 29~518 (1.05\%) 
of the 2~804~492 wordforms contained in our FST are stress-ambiguous. \rr{These 
stress-ambiguous wordforms have an average of 2.01 possible stress positions,
so a perfect stress placement tool has a 50/50 chance of getting them right.
This means that a perfect stress placement tool could achieve 99.48\% on their
task.} In our corpus of running text, at least 7.5\% of the tokens are stress-
ambiguous. Stress ambiguity is
therefore 7 times more prevalent in our corpus of running text than it is in 
our wordform dictionary.
Since the task of wordstress placement is virtually always performed on
running text, it seems prudent to make use of surrounding information.
The experiment described in this paper demonstrates that a Constraint Grammar
can effectively improve the accuracy of a stress placement system without 
significantly raising the error rate. Our Russian Constraint Grammar is
under continual development, so we expect higher accuracy in the future.

@@

We have demonstrated the effectiveness of using a Constraint Grammar to improve
the results of a Russian wordstress placement task in running text by resolving
42\% of the ambiguity in our gold corpus. We showed
that stress ambiguity is 7 times more prevalent in our corpus of running text
than it is in our lexicon, suggesting the importance of context-based
disambiguation for this task.  As with any 
lexicon- and rule-based system, the lexicon and rules can be expanded and 
improved, but our initial results are promising, especially considering the short 
timespan over which they were developed.

>>>>

\subsection{Background and task definition} \label{background}

@@

\subsection{Stress corpus}

@@

\section{Automatic stress placement}

State-of-the-art morphological analysis in Russian is based on finite-state
technology\cite{Nozhov-03,Segalovich-03}. To our knowledge, no existing 
open-source resources are available for analyzing and generating 
stressed wordforms. Therefore, we developed free and open-source finite-state 
tools capable of analyzing and generating stressed wordforms, based on the
well-known \emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}.
A Finite-State Transducer\footnote{Using two-level 
morphology \cite{Koskenniemi-84}, implemented in xfst \cite{Beesley.Karttunen-03} and hfst \cite{hfst-11}} (FST) generates all possible 
morphosyntactic readings of each wordform, and a constraint 
grammar\footnote{Implemented using vislcg3 constraint grammar parser (http://beta.visl.sdu.dk/cg3.html).}
\cite{Karlsson-90,Karlsson.Voutilainen.ea-95} then removes
some readings based on syntactic context. The ultimate success of our 
stress placement system depends on the performance of the Constraint Grammar. 
Ideally, the Constraint Grammar would successfully remove all but the correct 
reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

@@

The {\small {\tt guessFreq}} 
approach is to select the reading that is most frequent, with frequency data
taken from a separate hand-disambiguated corpus. If none of the readings are 
found in the corpus, then {\small {\tt guessFreq}} selects the reading
with the tag sequence (lemma removed) that is most frequent in our corpus. If the 
tag sequence is not found
in our frequency list, then {\small {\tt guessFreq}} backs off to the 
{\small {\tt guess}} algorithm. In the first column of
Table~\ref{tab:conditions}, {\small {\tt guessFreq}} selects \rus{к\'{о}сти} 
\emph{k\'{o}sti} because the tag sequence N-F-SG-GEN is more frequent than the
other alternatives. Note 
that for tokens with stress-irrelevant ambiguity (e.g. the second column of
Table~\ref{tab:conditions}), {\small {\tt guess}} and 
{\small {\tt guessFreq}} produce the same result as the {\small {\tt safe}} 
method.

@@

We noted in Section~\ref{background} that \newcite{hall_sproat_russian_2013}
achieved 98.7\% accuracy on stress placement for individual wordforms 
in a list (i.e. \emph{without} syntax). This result is 8.63\% higher than
{\small {\tt noCGsafe}}, but it is also a fundamentally different task.
Based on the surface 
forms in our FST -- which is based on the same dictionary used for 
\newcite{hall_sproat_russian_2013} -- 
we calculate that only 29~518 (1.05\%) 
of the 2~804~492 wordforms contained in our FST are stress-ambiguous. \rr{These 
stress-ambiguous wordforms have an average of 2.01 possible stress positions,
so a perfect stress placement tool has a 50/50 chance of getting them right.
This means that a perfect stress placement tool could achieve 99.48\% on their
task.} In our corpus of running text, at least 7.5\% of the tokens are stress-
ambiguous. Stress ambiguity is
therefore seven times more prevalent in our corpus of running text than it is in 
our wordform dictionary.
Since the task of wordstress placement is virtually always performed on
running text, it seems prudent to make use of surrounding information.
The experiment described in this paper demonstrates that a Constraint Grammar
can effectively improve the accuracy of a stress placement system without 
significantly raising the error rate. Our Russian Constraint Grammar is
under continual development, so we expect higher accuracy in the future.

@@

We have demonstrated the effectiveness of using a Constraint Grammar to improve
the results of a Russian wordstress placement task in running text by resolving
42\% of the ambiguity in our gold corpus. We showed
that stress ambiguity is seven times more prevalent in our corpus of running text
than it is in our lexicon, suggesting the importance of context-based
disambiguation for this task.  As with any 
lexicon- and rule-based system, the lexicon and rules can be expanded and 
improved, but our initial results are promising, especially considering the short 
timespan over which they were developed.
